Back-integration likelihood

### 2014-01-20 text/notes/likelihood/likelihood.tex
changed some notation

% bold X and D
\newcommand{\D}{{\bf D}}
\newcommand{\W}{{\bf W}}
\newcommand{\X}{{\bf X}}
\newcommand{\bSigma}{{\bf \Sigma}}
\newcommand{\bsigma}{\boldsymbol\sigma}

@@

Starting with notation, take ${\bf D}^{(i)}$ to be the observed, heliocentric 6D position of the $i$th star -- e.g., the measured position on the sky, ($l$, $b$), distance, $D$, proper motions, ($\mu_l$, $\mu_b$), and radial velocity, $v_r$ -- and ${\bf W}^{(i)}$ to be the corresponding \emph{true} 6D position of the same star in the same coordinate and reference frame. We assume that this star was once part of a progenitor system -- e.g., satellite galaxy -- presently at observed position ${\bf D}_s$ (with true position, ${\bf W}_s$ and shape described by a covariance matrix, $\bSigma_s$) that is tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The true 6D \emph{spherical, heliocentric} positions of the stars, $\W^{(i)}$, and satellite, $\W_s$, have corresponding 6D \emph{cartesian, Galactocentric} coordinate vectors $\X^{(i)}$ and $\X_s$. We then model the measured position of a star today as the result of sampling from a time-dependent 6D distribution -- the progenitor system -- at some time $\tau^{(i)}$ in the past, evolved to present day in the parent potential, neglecting the effect of self-interaction with the halo of the progenitor. We model the progenitor as a 6D Gaussian with mean vector given by the position of the satellite at any given time, ${\bf W}_s(t)$, and with shape defined by a time-dependent covariance matrix $\bSigma_s(t)$ (the subscript $s$ refers to satellite or progenitor). With these assumptions we can write the posterior for a single star as:
\begin{align}
	p(\Phi, \W^{(i)}, \W_s, \bSigma_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
		\frac{1}{Z} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\W^{(i)} \given \W_s, \bSigma_s, \tau^{(i)}, \Phi)\\
	p(\D^{(i)} \given \W^{(i)}) &= \mathcal{N}(\W^{(i)} \given \D^{(i)}, \bsigma^{(i)})\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bsigma_s)\\
	p(\W^{(i)} \given \W_s, \bSigma_s, \tau^{(i)}, \Phi) &= \mathcal{N}(\widetilde{\W}^{(i)}(\tau^{(i)}) \given \widetilde{\W}_s(\tau^{(i)}), \bSigma_s(\tau^{(i)}))
\end{align}
where the vectors $\bsigma^{2}$ and $\bsigma_s^2$ represent the (assumed) Gaussian, uncorrelated observational variances on the observed 6D position of star and the satellite, respectively.
%, and the tildes denote a conversion from spherical, heliocentric to cartesian, Galactocentric coordinates. 
%Thus, 
%\begin{align}
%	p(\Phi, \W, \W_s, \bSigma_s, \tau \given \D, \D_s) &= \prod_i p(\Phi, \W^{(i)}, \W_s, \bSigma_s, \tau^{(i)} \given \D^{(i)}, \D_s)\\
%	p &= \prod_i p^{(i)}\\
%	\ln p &= \sum_i \ln p^{(i)}\\
%	\ln p = \sum_i [\ln\mathcal{N}(\W^{(i)} \given \D^{(i)}, \bsigma^{(i)}) &+ \ln\mathcal{N}(\W_s \given \D_s, \bsigma_s) + \ln\mathcal{N}(\W^{(i)}(\tau^{(i)}) \given \W_s(\tau^{(i)}), \bSigma_s(\tau^{(i)}))]
%\end{align}

>>>>

% bold X and D
\newcommand{\D}{{\bf D}}
\newcommand{\W}{{\bf W}}
\newcommand{\X}{{\bf X}}
\newcommand{\V}{{\bf V}}
\newcommand{\bSigma}{{\bf \Sigma}}
\newcommand{\bsigma}{\boldsymbol\sigma}

@@

Starting with notation, take ${\bf D}^{(i)}$ to be the observed, heliocentric 6D position of the $i$th star -- e.g., the measured position on the sky, ($l$, $b$), distance, $D$, proper motions, ($\mu_l$, $\mu_b$), and radial velocity, $v_r$ -- and ${\bf W}^{(i)}$ to be the corresponding \emph{true}, 6D position of the same star in the same coordinate and reference frame. We assume that this star was once part of a progenitor system -- e.g., satellite galaxy -- tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The satellite is observed to be at heliocentric 6D position ${\bf D}_s$, with true position ${\bf W}_s$, and shape described by a covariance matrix, $\V_s$ (the subscript $s$ refers to the satellite). The true, 6D spherical, heliocentric positions of the stars and satellite -- $\W^{(i)}$ and $\W_s$ -- have corresponding 6D cartesian, Galactocentric coordinate vectors $\X^{(i)}$ and $\X_s$.

We model the observed, present-day position of a star as the result of sampling from a time-dependent 6D distribution -- the progenitor system -- at some time $\tau^{(i)}$ in the past, evolved to present day in the parent potential, neglecting the effect of self-interaction with the halo of the progenitor. We model the progenitor as a 6D Gaussian with mean vector given by the position of the satellite at any given time, ${\bf W}_s(t)$, and with shape defined by the time-dependent covariance matrix $\V_s(t)$. With these assumptions we can write the joint posterior probability for the observed position of a star, properties of the disrupting satellite, and parent potential given the data as:
\begin{align}
	%p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
	%	\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi)
	p(\Phi, \W^{(i)}, \W_s, \V_s, \tau^{(i)} \given \D^{(i)}, \D_s) &= 
		\frac{1}{\mathcal{Z}} p(\D^{(i)} \given \W^{(i)}) p(\D_s \given \W_s) p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian:
\begin{align}
	p(\D^{(i)} \given \W^{(i)}) &= \mathcal{N}(\W^{(i)} \given \D^{(i)}, \bSigma^{(i)})\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bSigma_s)
\end{align}
where the covariance matrices $\bSigma^{(i)}$ and $\bSigma_s$ represent the observational variances on the observed 6D position of star and the satellite, respectively. [...] hmm...
\begin{align}
	p(\D^{(i)} \given \D_s, \V_s, \tau^{(i)}, \Phi) &= p(\W^{(i)} \given \W_s, \V_s, \tau^{(i)}, \Phi) \times \det{\bf J}^{(i)}\\
	&= \left(\mathcal{N}(\W^{(i)} \given \W_s, \V_s) \times \det{\bf J}^{(i)}\right)\rvert_{t=\tau^{(i)}}
\end{align}

%, and the tildes denote a conversion from spherical, heliocentric to cartesian, Galactocentric coordinates. 
%Thus, 
%\begin{align}
%	p(\Phi, \W, \W_s, \bSigma_s, \tau \given \D, \D_s) &= \prod_i p(\Phi, \W^{(i)}, \W_s, \bSigma_s, \tau^{(i)} \given \D^{(i)}, \D_s)\\
%	p &= \prod_i p^{(i)}\\
%	\ln p &= \sum_i \ln p^{(i)}\\
%	\ln p = \sum_i [\ln\mathcal{N}(\W^{(i)} \given \D^{(i)}, \bsigma^{(i)}) &+ \ln\mathcal{N}(\W_s \given \D_s, \bsigma_s) + \ln\mathcal{N}(\W^{(i)}(\tau^{(i)}) \given \W_s(\tau^{(i)}), \bSigma_s(\tau^{(i)}))]
%\end{align}
